---
title: "Tree mapping"
date: '`r format(Sys.time(), usetz = TRUE)`'
output:
  bookdown::pdf_document2:
    toc: no
    extra_dependencies: ["float"]
csl: "assets/palaeontology.csl"
link-citations: yes
bibliography:
  - assets/REFERENCES.bib
---

\renewcommand{\figurename}{Supplementary Figure}

```{r initialize-variables, echo=FALSE, message=FALSE}
# Set preferred distance method.  Choose from "CID" or "Quartet".
# See Smith (2020, https://doi.org/10.1093/bioinformatics/btaa614) and
# Smith (2022, https://doi.org/10.1093/sysbio/syab100) for guidance.
distMethod <- "CID"

# What taxon / taxa should be used to root the tree?
outgroup <- "Tubiluchus"

# Paths to results directories
treeSearchDir <- "../TreeSearch" # Must contain the input matrix file
mbDir <- "../MrBayes"            # Location of MrBayes results, if they exist
rbDir <- "../RevBayes"           # Location of RevBayes results, if they exist

# Run time scales with the square of the number of trees visualized.
# How many trees should be sampled from each set of parsimony results?
nMapPars <- 24
# How many trees should be sampled from each run of each Bayesian analysis?
nMapBayes <- 24
# How many trees should be sampled when computing posterior probabilities?
ppSample <- 500
# What proportion of Bayesian trees should be discarded as burnin? (0.0-1.0)
burninF <- 0.10

# What k values from implied weight parsimony should be visualized?
# If NULL, all values will be used.
concavities <- NULL
# Trees with these low values of K will be visualized separately.
# Set to NULL to visualize all implied weighting trees together.
lowK <- c(3, 6)

# Set output parameters
pageHeight <- (254 / 25.4)
pageWidth <- 176 / 25.4
colWidth <- 84 / 25.4
midWidth <- 120 / 25.4

# Define colour scale for concordance values
concHead <- 51
concScale <- c(rep("#FFFFFF", concHead),
               hcl.colors(101 - concHead, "plasma", rev = TRUE))
concLegScale <- concScale[-seq_len(concHead)]
```

```{r set-up-session, echo=FALSE, message=FALSE}
# Load libraries and functions
library("ape")
library("phangorn")
library("TreeTools", quietly = TRUE)
library("TreeSearch")
#library("cli")
source(paste0(treeSearchDir, "/common.R"))

dataset <- LatestMatrix(treeSearchDir, full.names = FALSE)
if (is.na(dataset)) {
  stop("Dataset could not be found. Check a Nexus file exists in ",
     treeSearchDir)
}
```

```{r load-data, echo=FALSE}
#| eval = !is.na(dataset)
dat <- ReadAsPhyDat(LatestMatrix(treeSearchDir, full.names = TRUE))
latest <- LatestMatrix(treeSearchDir)
dat <- ReadAsPhyDat(latest)
treeFiles <- list.files(
  path = treeSearchDir,
  pattern = paste0(".+_", sub("^.*/", "", latest), ".trees"),
  full.names = TRUE
)
if (is.null(concavities)) {
  concavities <- sort(KValue(treeFiles))
  concavities <- concavities[is.finite(concavities)]
}
```

```{r output-settings, echo = FALSE}
set.seed(1)
knitr::opts_chunk$set(fig.pos = "H", out.extra = "", fig.width = 6.8)
```

```{r load-trees, echo = FALSE}
tsPrefix <- c("ew", paste0("iw", concavities))
tsFiles <- paste0(treeSearchDir, "/", tsPrefix, "_", dataset, ".trees")
tsPrefix <- tsPrefix[file.exists(tsFiles)]
tsFiles <- tsFiles[file.exists(tsFiles)]

escapeDataset <- gsub("([\\-\\.])", "\\\\\\1", dataset)
rbFiles <- list.files(rbDir, paste0("^", escapeDataset, "\\.mkv_run_\\d+\\.trees$"),
                      full.names = TRUE)
mbFiles <- list.files(mbDir, paste0("^", escapeDataset, "\\.run\\d+\\.t$"),
                      full.names = TRUE)

tsTrees <- if (length(tsFiles)) {
  setNames(lapply(tsFiles, read.nexus), paste0("ts", tsPrefix))
} else {
  NULL
}
lowKNames <- paste0("tsiw", lowK)
highKTrees <- tsTrees[setdiff(names(tsTrees), c(lowKNames, "tsew"))]
lowKTrees <- tsTrees[intersect(names(tsTrees), lowKNames)]

# Determine whether there's anything to analyse in subsequent chunks
evalHighK <- !is.null(highKTrees) && length(highKTrees) > 0
evalLowK <- !is.null(lowKTrees) && length(lowKTrees) > 0
evalEW <- !is.null(tsTrees[["tsew"]])

rbTrees <- if (length(rbFiles)) {
  lapply(rbFiles,
    function(rbFile) {
      treeTable <- read.table(rbFile, header = TRUE, sep = "\t")
      read.tree(text = gsub(" ", "_", fixed = TRUE, treeTable[["phylogeny"]])) |>
        setNames(paste0("it", treeTable[["Iteration"]]))
    }) |>
  RootTree(outgroup) |>
   setNames(paste0("rb", seq_along(rbFiles)))
} else {
  NULL
}

mbTrees <- if (length(mbFiles)) {
  lapply(
    mbFiles,
    function (mbFile) tryCatch(
      read.nexus(mbFile),
      error = function (err) if (err$message == "NA/NaN argument") {
        # Unterminated tree block, perhaps because a search is ongoing
        withEnd <- tempfile()
        on.exit(unlink(withEnd))
        writeLines(c(readLines(mbFile), "\nEND;"), withEnd)
        read.nexus(withEnd)
      })) |> setNames(paste0("mb", seq_along(mbFiles)))
} else {
  NULL
}

# Remove burnin
mbBurnOff <- lapply(mbTrees, function(trees) {
  nTree <- length(trees)
  trees[((nTree * burninF) + 1):nTree]
})
nMb <- sum(lengths(mbBurnOff))
mbPosterior <- do.call(c, mbBurnOff)[
  seq.int(1, nMb, length.out = min(nMb, ppSample))] |> 
  RootTree(outgroup)

rbBurnOff <- lapply(rbTrees, function(trees) {
  nTree <- length(trees)
  trees[(1 + (nTree * burninF)):nTree]
})
nRb <- sum(lengths(rbBurnOff))
rbPosterior <- do.call(c, rbBurnOff)[
  seq.int(1, nRb, length.out = min(nRb, ppSample))] |> 
  RootTree(outgroup)


Sample <- function (x, n) {
  if (length(x) > n) {
    sample(x, n)
  } else {
    x
  }
}

set.seed(1) # Consistent sampling
#tntSample <- lapply(tntTrees, Sample, nMapPars)
tsSample <- lapply(tsTrees, Sample, nMapPars)
mbSample <- lapply(mbBurnOff, Sample, nMapBayes)
rbSample <- lapply(rbBurnOff, Sample, nMapBayes)
```

# Summary of phylogenetic results

Phylogenetic inference was conducted using Bayesian and parsimony methods.
Whereas there are many reasons to prefer implied weights to equal weights
parsimony [@Goloboff2018; @Smith2019; @Ezcurra2024], neither Bayesian nor
parsimony approaches consistently recover 'better' trees once precision and
accuracy are taken into account [@Smith2019].
We present results of multiple methods, on the basis that relationships that are
consistently recovered are less likely to represent methodological artefacts,
and so more likely to represent genuine evolutionary signal in the data.

To summarize our results, we present consensus trees constructed after identifying rogue taxa using method H1 from @SmithRogue.
Rogue taxa are wildcards whose position varies from tree to tree due to ambiguity or conflict in their character data [@Kearney2002]; removal of such taxa from input trees reveals consistent relationships between the remaining taxa in a consensus summary tree that would otherwise be masked by the uncertain position of the rogues [@Wilkinson1994;@Wilkinson1996hb;@Wilkinson1996mbe].
Splits in Bayesian trees are labelled with their posterior probability; in parsimony trees, splits are labelled with their quartet concordance [@Minh2020mbe].

## Bayesian analysis

```{r install-treess, echo = FALSE}
#| eval = (!is.null(mbPosterior) || !is.null(rbPosterior)) &&
#|  !requireNamespace("treess", quietly = TRUE)

message("Installing packages necessary to calculate tree ESS")
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
devtools::install_github("afmagee/treess")
```
```{r bayes-functions, echo = FALSE}
BayesCons <- function(posterior) {
  cons <- SortTree(RootTree(consensus(posterior, p = 0.5), outgroup))
  plot(cons)
  splitFreqs <- SplitFrequency(cons, posterior) / length(posterior)
  tooBright <- 9
  LabelSplits(
    cons,
    round(splitFreqs * 100),
    col = SupportColour(
      splitFreqs,
      scale = c(rep("white", 50 - tooBright),
                hcl.colors(51 + tooBright, "plasma", rev = TRUE))
      ),
    frame = "none",
    pos = 1,
    cex = 0.8
  )
  
  PlotTools::SpectrumLegend(
    "bottomright",
    legend = c(100, 75, 50),
    title = "PP / %",
    palette = hcl.colors(51 + tooBright, "plasma",
                         rev = TRUE)[-seq_len(tooBright)],
    xpd = NA, bty = "n"
  )
}

BayesRogues <- function(posterior, rogueTaxa) {
  cons <- SortTree(RootTree(ConsensusWithout(posterior, rogueTaxa, p = 0.5), outgroup))
  plot(cons)
  splitFreqs <- SplitFrequency(cons, posterior) / length(posterior)
  tooBright <- 9
  LabelSplits(
    cons,
    round(splitFreqs * 100),
    col = SupportColour(
      splitFreqs,
      scale = c(rep("white", 50 - tooBright),
                hcl.colors(51 + tooBright, "plasma", rev = TRUE))
      ),
    frame = "none",
    pos = 1,
    cex = 0.8
  )
  
  legend("bottomleft", bty = "n",
         lty = 2, text.font = 3, cex = 0.8,
         title = "Rogue taxa", title.font = 1,
         rogueTaxa)
  
  PlotTools::SpectrumLegend(
    "bottomright",
    legend = c(100, 75, 50),
    title = "PP / %",
    palette = hcl.colors(51 + tooBright, "plasma",
                         rev = TRUE)[-seq_len(tooBright)],
    xpd = NA, bty = "n"
  )
}

BayesESS <- function(posterior, analysisName) {
  ess <- treess::treess(list(posterior), TreeDist::RobinsonFoulds,
                        methods = c("frechetCorrelationESS", "medianPseudoESS"))[[1]]
  cat(paste("The posterior sample of", length(posterior), analysisName,
            "trees has Frechet Correlation ESS of",
            signif(ess[["frechetCorrelationESS"]], 4),
            "and Median Pseudo ESS of",
            signif(ess[["medianPseudoESS"]], 4),
            "[@Lanfear2016;@Magee2024]"
            ))
}
```

```{r mrbayes-tree-ess, echo = FALSE, results = "asis"}
#| eval = !is.null(mbPosterior) && requireNamespace("treess", quietly = TRUE)
BayesESS(mbPosterior, "MrBayes")
```
```{r plot-mrbayes, echo = FALSE}
#| eval = !is.null(mbPosterior),
#| fig.height = 6.4,
#| fig.cap = "Majority rule consensus of Bayesian trees from MrBayes analysis;
#|   splits labelled with posterior probabilities."
par(mar = rep(0, 4), cex = 0.8)
BayesCons(mbPosterior)
```

```{r mrbayes-rogues, echo = FALSE}
#| eval = !is.null(mbPosterior),
#| message = FALSE,
#| results = "hide"
set.seed(1) # Ensure consistent result from heuristic search
mbRogues <- Rogue::QuickRogue(mbPosterior)
```
```{r plot-mrbayes-rogues, echo = FALSE}
#| eval = !is.null(mbPosterior) && nrow(mbRogues) > 1,
#| fig.height = 6.4,
#| fig.cap = "Majority rule consensus of Bayesian trees from MrBayes analysis,
#|   omitting rogue taxa;
#|   splits labelled with posterior probabilities."
par(mar = rep(0, 4), cex = 0.8)
BayesRogues(mbPosterior, mbRogues[-1, "taxon"])
```


```{r revbayes-tree-ess, echo = FALSE, , results = "asis"}
#| eval = !is.null(rbPosterior) && requireNamespace("treess", quietly = TRUE)
BayesESS(rbPosterior, "RevBayes")
```
```{r plot-revbayes, echo = FALSE}
#| eval = !is.null(rbPosterior),
#| fig.height = 6.4,
#| fig.cap = "Majority rule consensus of Bayesian trees from RevBayes analysis;
#|   splits labelled with posterior probabilities."
par(mar = rep(0, 4), cex = 0.8)
BayesCons(rbPosterior)
```

```{r revbayes-rogues, echo = FALSE}
#| eval = !is.null(rbPosterior),
#| message = FALSE,
#| results = "hide"
set.seed(1) # Ensure consistent result from heuristic search
rbRogues <- Rogue::QuickRogue(rbPosterior)
```
```{r plot-revbayes-rogues, echo = FALSE}
#| eval = !is.null(rbPosterior) && nrow(rbRogues) > 1,
#| fig.height = 6.4,
#| fig.cap = "Majority rule consensus of Bayesian trees from RevBayes analysis,
#|   omitting rogue taxa;
#|   splits labelled with posterior probabilities."
par(mar = rep(0, 4), cex = 0.8)
BayesRogues(rbPosterior, rbRogues[-1, "taxon"])
```

## Maximum parsimony

```{r pars-highk, echo = FALSE}
#| eval = evalHighK
trees <- RootTree(do.call(c, highKTrees), outgroup)
```

```{r highk-rogues, echo = FALSE}
#| eval = evalHighK,
#| message = FALSE,
#| results = "hide"
set.seed(1) # Ensure consistent result from heuristic search
highKRogues <- Rogue::QuickRogue(trees, p = 1)
highKRogues
```
```{r plot-pars-cons, echo = FALSE}
PlotParsCons <- function(trees) {
  cons <- SortTree(RootTree(consensus(trees), outgroup))
  plot(cons)
  concordance <- QuartetConcordance(cons, dat)
  LabelSplits(
    tree = cons,
    labels = round(concordance * 100),
    col = SupportColor(concordance, scale = concScale),
    frame = "none",
    pos = 3,
    cex = 0.8
  )
  PlotTools::SpectrumLegend(
    "bottomright",
    legend = c(100, 75, 50),
    title = "Concordance / %",
    palette = concLegScale,
    xpd = NA, bty = "n"
  )
}

PlotParsRogues <- function(trees, rogueTaxa) {
  if (length(rogueTaxa) == 1) {
    rp <- RoguePlot(trees, rogueTaxa, sort = TRUE)
    cons <- rp[["cons"]]
    PlotTools::SpectrumLegend(
      "bottomleft", 
      palette = colorRampPalette(c(par("fg"), "#009E73"), space = "Lab")(100),
      legend = rp[["legendLabels"]],
      cex = 0.8, bty = "n", title = paste("Position of", rogueTaxa, "in")
    )
  } else {
    cons <- SortTree(RootTree(ConsensusWithout(trees, rogueTaxa),
                              setdiff(outgroup, rogueTaxa)))
    plot(cons)
    legend("bottomleft", bty = "n",
           lty = 2, text.font = 3, cex = 0.8,
           title = "Rogue taxa", title.font = 1,
           rogueTaxa)
  }
  concordance <- QuartetConcordance(cons, dat)
  LabelSplits(
    tree = cons,
    labels = round(concordance * 100),
    col = SupportColor(concordance, scale = concScale),
    frame = "none",
    pos = 3,
    cex = 0.8
  )
  PlotTools::SpectrumLegend(
    "bottomright",
    legend = c(100, 75, 50),
    title = "Concordance / %",
    palette = concLegScale,
    xpd = NA, bty = "n"
  )
}
```
```{r plot-pars-highK, echo = FALSE}
#| eval = evalHighK,
#| message = FALSE,
#| fig.height = 6.4,
#| fig.cap = "(ref:pars-highK-cap)"

par(mar = rep(0, 4), cex = 0.8)
PlotParsCons(trees)
```
(ref:pars-highK-cap) Strict consensus of results for implied weights parsimony with high concavity values (`r paste(sort(setdiff(concavities, lowK)), collapse = ", ")`).  Splits are labelled with their quartet concordance [@Minh2020mbe].

```{r plot-pars-highK-rogue, echo = FALSE}
#| eval = evalHighK && nrow(highKRogues) > 1,
#| message = FALSE,
#| fig.height = 6.4,
#| fig.cap = "(ref:pars-highK-rogue-cap)"

par(mar = rep(0, 4), cex = 0.8)
PlotParsRogues(trees, highKRogues[-1, "taxon"])
```

(ref:pars-highK-rogue-cap) Strict consensus of results for implied weights parsimony with high concavity values (`r paste(sort(setdiff(concavities, lowK)), collapse = ", ")`), after removal of rogue taxa. Splits are labelled with their quartet concordance [@Minh2020mbe].


```{r lowK-rogues, echo = FALSE}
#| eval = evalLowK,
#| message = FALSE,
#| results = "hide"

set.seed(1) # Consistent result from heuristic search
trees <- RootTree(do.call(c, lowKTrees), outgroup)
lowKRogues <- Rogue::QuickRogue(trees, p = 1)
lowKRogues
```
```{r plot-pars-lowK, echo = FALSE}
#| eval = evalLowK,
#| message = FALSE,
#| fig.height = 6.4,
#| fig.cap = "(ref:pars-lowK-cap)"

par(mar = rep(0, 4), cex = 0.8)
PlotParsCons(trees)
```
(ref:pars-lowK-cap) Strict consensus of results for implied weights parsimony with low concavity values (`r paste(sort(lowK), collapse = ", ")`).  Splits are labelled with their quartet concordance [@Minh2020mbe].

```{r plot-pars-lowK-rogue, echo = FALSE}
#| eval = evalLowK && nrow(lowKRogues) > 1,
#| message = FALSE,
#| fig.height = 6.4,
#| fig.cap = "(ref:pars-lowK-rogue-cap)"

par(mar = rep(0, 4), cex = 0.8)
PlotParsRogues(trees, lowKRogues[-1, "taxon"])
```

(ref:pars-lowK-rogue-cap) Strict consensus of results for implied weights parsimony with low concavity values (`r paste(lowK, collapse = ", ")`), after removal of rogue taxa. Splits are labelled with their quartet concordance [@Minh2020mbe].

```{r ew-rogues, echo = FALSE}
#| eval = evalEW,
#| message = FALSE,
#| results = "hide"

set.seed(1) # Ensure consistent result from heuristic search
trees <- RootTree(tsTrees[["tsew"]], outgroup)
ewRogues <- Rogue::QuickRogue(trees, p = 1)
ewRogues
```

```{r plot-pars-ew, echo = FALSE}
#| eval = evalEW,
#| message = FALSE,
#| fig.height = 6.4,
#| fig.cap = "(ref:pars-ew-cap)"

par(mar = rep(0, 4), cex = 0.8)
PlotParsCons(trees)
```
(ref:pars-ew-cap) Strict consensus of results for equal weights parsimony, after removal of rogue taxa.  Splits are labelled with their quartet concordance [@Minh2020mbe].

```{r plot-pars-ew-rogue, echo = FALSE}
#| eval = evalEW && nrow(ewRogues) > 1,
#| message = FALSE,
#| fig.height = 6.4,
#| fig.cap = "(ref:pars-ew-rogue-cap)"

par(mar = rep(0, 4), cex = 0.8)
PlotParsRogues(trees, ewRogues[-1, "taxon"])
```

(ref:pars-ew-rogue-cap) Strict consensus of results for equal weights parsimony, after removal of rogue taxa.  Splits are labelled with their quartet concordance [@Minh2020mbe].



<!-- compute distances for clustering and mapping -->

```{r tree-distances, echo = FALSE}
#| message = FALSE,
#| results = "hide"

library("TreeDist")

pal <- hcl.colors(4, palette = "Dynamic")

allTrees <- c(
  do.call(c, tsSample),
  do.call(c, mbSample),
  do.call(c, rbSample)
)

lengths <- c(
  vapply(tsSample, length, 1),
  vapply(mbSample, length, 1),
  vapply(rbSample, length, 1)
)

names <- rep(names(lengths), lengths)

if (is.null(allTrees)) {
  stop("No trees have been loaded; check that results exist")
}

distances <- switch(
  distMethod,
  "Quartet" = as.dist(Quartet::QuartetDivergence(
    Quartet::ManyToManyQuartetAgreement(allTrees),
    similarity = FALSE)),
  "CID" = ClusteringInfoDist(allTrees)
)
```

```{r map-distances, echo = FALSE}
set.seed(1) # Consistent mapping output
mapping <- cmdscale(distances, k = 10)
mstEnds <- MSTEdges(distances)
```


# Cluster analysis

```{r initialize-constants, echo = FALSE}
blackGap <- 4L

tsPal <- setNames(
  colorRampPalette(c(pal[[1]], "black"), space = "Lab")(
    length(concavities) + 1 + blackGap
  ),
  paste0("ts", c(if (length(concavities) > 0) paste0("iw", sort(concavities)),
                 "ew", rep(NA, blackGap)))
)

mbPal <- setNames(colorRampPalette(c(pal[[2]], "black"),
                                    space = "Lab")(length(mbFiles) + blackGap),
                   paste0("mb", c(seq_along(mbFiles), rep(NA, blackGap))))

rbPal <- setNames(colorRampPalette(c(pal[[3]], "black"),
                                    space = "Lab")(length(rbFiles) + blackGap),
                   paste0("rb", c(seq_along(rbFiles), rep(NA, blackGap))))

cols <- c(tsPal, mbPal, rbPal)[names]
```
```{r calculate-clusterings, echo = FALSE}
possibleClusters <- 2:(2 * length(concavities) + 4)

pamClusters <- lapply(possibleClusters, function (k) cluster::pam(distances, k = k))
pamSils <- vapply(pamClusters, function (pamCluster) {
  mean(cluster::silhouette(pamCluster)[, 3])
}, double(1))

bestPam <- which.max(pamSils)
pamSil <- pamSils[bestPam]
pamCluster <- pamClusters[[bestPam]]$cluster

hTree <- protoclust::protoclust(distances)
hClusters <- lapply(possibleClusters, function (k) cutree(hTree, k = k))
hSils <- vapply(hClusters, function (hCluster) {
  mean(cluster::silhouette(hCluster, distances)[, 3])
}, double(1))


bestH <- which.max(hSils)
hSil <- hSils[bestH]
hCluster <- hClusters[[bestH]]


kClusters <- lapply(possibleClusters, function (k) KMeansPP(distances, k = k))
kSils <- vapply(kClusters, function (kCluster) {
  mean(cluster::silhouette(kCluster$cluster, distances)[, 3])
}, double(1))


bestK <- which.max(kSils)
kSil <- kSils[[bestK]]
kCluster <- kClusters[[bestK]]$cluster

dbClusters <- lapply(c(5, 10, 25, 50), function(pts) {
  tryCatch(dbscan::hdbscan(distances, minPts = pts),
           error = function(e) {
             structure(list("cluster_scores" = NA), class = "hdbscan")
           })
})
dbN <- vapply(dbClusters, function(clust) length(clust[["cluster_scores"]]),
              double(1))
dbClusters <- dbClusters[dbN > 1]
dbN <- dbN[dbN > 1]
distMat <- as.matrix(distances)
dbSils <- vapply(dbClusters, function(dbCluster) {
  cl <- dbCluster$cluster
  mean(cluster::silhouette(cl[cl != 0], distMat[cl != 0, cl != 0])[, 3])
}, double(1))
bestDB <- which.max(dbSils)
dbSil <- dbSils[[bestDB]]
dbCluster <- dbClusters[[bestDB]]$cluster
```


Clusters of trees were evaluated using partitioning around medoids [@Maechler2022],
hierarchical clustering with minimax linkage [@Bien2011; @Murtagh1983],
K-means++ clustering [@Arthur2007; @Hartigan1979; @TreeDist], and
hierarchical density based clustering [@Campello2015; @Hahsler2019; @Hahsler2023].
The clustering quality under each method, with 2–`r max(possibleClusters)` clusters, was compared using the silhouette coefficient [@Rousseeuw1987; @Maechler2022]; high silhouette coefficients distinguish the most clearly defined clusters.

```{r plot-cluster-summary, echo = FALSE, fig.cap = "Evaluation of optimal clustering"}
par(mar = c(4, 4, 0.1, 0.1), cex = 0.7)
plot(pamSils ~ possibleClusters,
     xlab = "Number of clusters", ylab = "Silhouette coefficient",
     ylim = range(c(pamSils, hSils, kSils)),
     frame.plot = FALSE)
points(hSils ~ possibleClusters, pch = 2, col = 2)
points(kSils ~ possibleClusters, pch = 3, col = 3)
points(dbSils ~ dbN, pch = 4, col = 4)
legend("topright", bty = "n",
       c("PAM", "Hierarchical", "K-means++", "HDBSCAN"),
       pch = 1:4, col = 1:4)

cluster <- switch(which.max(c(hSil, pamSil, kSil, dbSil)),
                  hCluster, pamCluster, kCluster, dbCluster)
```

```{r hcut, echo = FALSE, eval = FALSE}
class(hTree) <- "hclust"

par(mar = c(0, 0, 0, 0))
plot(hTree, labels = names, main = "", cex = 0.5)
points(seq_along(cols), rep(5, length(cols)), pch = 16,
       col = cols[hTree$order])
```


Plotting the consensus tree for each cluster can reveal resolution that is consistent within each cluster, but obscured in a single consensus of all topologies.

```{r cluster-consensus-trees, echo = FALSE}
#| fig.height = 6.4,
#| fig.cap = "(ref:clust-cons-cap)"

calcPP <- sum(lengths(c(mbPosterior, rbPosterior))) > 0
PlotGrid <- function(n) rep(ceiling(sqrt(n)), 2)
cl <- table(cluster[cluster > 0])
par(mfrow = PlotGrid(length(cl)), mar = rep(0.2, 4), cex = 0.7)

for (i in as.integer(names(cl[order(cl, decreasing = TRUE)]))) {
  trees <- allTrees[cluster == i]
  cons <- SortTree(RootTree(consensus(trees, p = 0.5), outgroup))
  plot(cons)
    
  legend("bottomleft",
         sprintf("Consensus of %g%% of trees",
                 signif(100 * length(trees) / length(allTrees), 3)),
         bty = "n")
  if (calcPP) {
    splitFreqs <- SplitFrequency(cons, c(mbPosterior, rbPosterior)) / 
      length(c(mbPosterior, rbPosterior))
    tooBright <- 26
    LabelSplits(cons, round(splitFreqs * 100),
                col = SupportColour(
                  splitFreqs,
                  scale = hcl.colors(101 + tooBright, "plasma",
                                     rev = TRUE)[-seq_len(tooBright)]
                ),
                frame = "none", pos = 1, cex = 0.8)
  }
}

if (calcPP) {
  PlotTools::SpectrumLegend(
    "bottomright",
    legend = c(100, 75, 50, 25, 0),
    title = "PP / %",
    palette = hcl.colors(101 + tooBright, "plasma",
                         rev = TRUE)[-seq_len(tooBright)],
    xpd = NA, bty = "n"
  )
}
```

(ref:clust-cons-cap) 50% majority rule consensus trees for each cluster demonstrate the topographic changes underpinning the differences between clusters.

# Tree space mapping

To evaluate the consistency of phylogenetic results, we map the phylogenetic trees obtained by different methods using the Quartet distance [@Estabrook1985; @SmithSpace], using the R [@R] packages "Quartet" [@Quartet; @Sand2014] and "TreeDist" [@TreeDist].  Distances in the mapping correspond to the degree of difference between tree topologies; dashed lines depict the shortest path to connect all trees, depicting any distortion introduced by the mapping [@SmithSpace].

The mapping is conducted on a subsample of `r nMapPars` trees from each parsimony search, and `r nMapBayes` samples from the posterior tree set of each Bayesian run.


```{r plot-mapping, echo = FALSE}
#| fig.asp = 1,
#| fig.cap = "Two-dimensional mapping of phylogenetic results"

par(mar = rep(0, 4), cex = 0.9)

plot(mapping, ann = FALSE, axes = FALSE, frame.plot = FALSE, type = "n",
     asp = 1)

# Plot MST
invisible(apply(mstEnds, 1, function (segment)
  lines(mapping[segment, 1], mapping[segment, 2], col = "#e0e0e0", lty = 2)))

# Add points
points(mapping, col = cols, pch = ifelse(substr(names, 1, 1) == "t", 16, 1))

nPars <- length(concavities) + 1L
legend(
  "bottomright",
  title = "Parsimony",
  bty = "n",
  pch = 16,
  col = tsPal[seq_len(nPars)],
  c(paste("k =", concavities), "Equal weights"),
  cex = 0.75, pt.cex = 1.2
)

legend(
  "bottomleft",
  title = "Bayesian / Mk",
  bty = "n",
  pch = 1,
  col = c(mbPal[seq_along(mbFiles)], rbPal[seq_along(rbFiles)]),
  c(if (length(mbFiles) > 0) paste("MrBayes run", seq_along(mbFiles)),
    if (length(rbFiles) > 0) paste("RevBayes run", seq_along(rbFiles))),
  cex = 0.75, pt.cex = 1.2
)
```

The mapping shows whether Bayesian and parsimony trees occupy a similar region of tree space, and at which concavity constants.

```{r evaluate-mapping, echo = FALSE}
qual <- vapply(seq_len(ncol(mapping)), function (k) {
  newDist <- dist(mapping[, seq_len(k)])
  MappingQuality(distances, newDist, 10)
}, double(4))

nDim <- max(2, which.max(qual["TxC", ] > 0.9))
```

The quality of the two-dimensional mapping (trustworthiness = `r signif(qual["Trustworthiness", 2], 3)`; continuity = `r signif(qual["Continuity", 2], 3)`), allows an evaluation of whether all aspects of the distances between pairs of trees are faithfully depicted. `r nDim` dimensions are required to meet the recommendation [@SmithSpace] of trustworthiness × continuity \> 0.9.


(ref:treespace-cap) The structure of tree space in higher dimensions is shown using a `r nDim`-dimensional tree space mapping, with clusters numbered and their convex hulls marked.

```{r cluster-plot, echo = FALSE}
#| fig.asp = 1, 
#| fig.cap = "(ref:treespace-cap)"

plotSeq <- matrix(0, nDim, nDim)
plotSeq[upper.tri(plotSeq)] <- seq_len(nDim * (nDim - 1) / 2)
plotSeq <- t(plotSeq[-nDim, -1])
if (nDim > 2) plotSeq[1, nDim - 1L] <- (nDim * (nDim - 1) / 2) + 1

layout(plotSeq)
par(mar = rep(0.1, 4))

for (i in 2:nDim) for (j in seq_len(i - 1)) {
  # Set up blank plot
  plot(mapping[, j], mapping[, i], ann = FALSE, axes = FALSE,
       frame.plot = nDim > 2, type = "n", asp = 1,
       xlim = range(mapping), ylim = range(mapping))

  mtext(paste("Dimension", i), 2, line = -1, cex = 0.6)
  mtext(paste("Dimension", j), 3, line = -1, cex = 0.6)

  # Plot MST
  apply(mstEnds, 1, function (segment)
    lines(mapping[segment, j], mapping[segment, i], col = "#bbbbbb", lty = 1))

  # Add points
  points(mapping[, j], mapping[, i], pch = 16, col = cols)

  # Mark clusters
  for (clI in unique(cluster)) {
    inCluster <- cluster == clI
    clusterX <- mapping[inCluster, j]
    clusterY <- mapping[inCluster, i]
    hull <- chull(clusterX, clusterY)
    polygon(clusterX[hull], clusterY[hull], lty = 1, lwd = 2,
            border = "#54de25bb")
    text(mean(clusterX), mean(clusterY), clI, col = "#54de25bb", font = 2)
  }
}

```

# References
<!-- References cited with @identifiers will be inserted here on Knit. -->
